A Makefile for Developing Containerized \LaTeX\ Technical Documents

### 2021-08-31 ms.tex
init

\begin{document}
\DTLloaddb{keys_values}{tmp/keys-values.csv}

@@

\begin{abstract}
	We propose a Makefile for developing containerized \LaTeX\ technical documents.
	The Makefile allows the author to execute the code that generates variables, tables and figures (results), which are then used during the \LaTeX\ compilation, to produce either the draft (fast) or full (slow) version of the document.
	We also present various utilities that aid in automating the results generation and improve the reproducibility of the document.
	We release an open source repository of a template that uses the Makefile and demonstrate its use by developing this paper.
\end{abstract}

@@

Although \textit{literate programming} provides many advantages for demonstrative projects and immediate visual feedback during experimentation, its `dual nature' constrains authors in using specifically designed text editors and/or source code formats thus leading to `app/vendor lock-in` situations.
Moreover, the overlap between natural text and results code makes \textit{literate programming} difficult to debug and version control.

We propose a Makefile for developing containerized \LaTeX\ technical documents.
By using portable technologies that withstood the test of time, we ensure the stability and portability of the proposed Makefile.
We write this paper using a template that consists of the Makefile and choose Docker as the container technology and the Python programming language for generating the results.

@@

\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[%
			grow via three points={one child at (0.5,-0.7) and two children at (0.5,-0.7) and (0.5,-1.4)},
			edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
			\node[draw](document){document/}
			child{node[draw, anchor=west](makefile){Makefile}}
			child{node[draw, anchor=west, fill=green!20](tmp){tmp/}}
			child{node[draw, anchor=west, fill=red!20](tex){ms.tex}}
			child{node[draw, anchor=west, fill=red!20](bib){ms.bib}}
			child{node[draw, anchor=west, fill=blue!20](py){main.py}}
			child{node[draw, anchor=west, fill=blue!20](containerfile){Containerfile}}
			child{node[draw, anchor=west, fill=blue!20](requirements){requirements.txt}};
			\begin{scope}[on background layer]
				\node[draw, rounded corners, dashed, fill=gray!20, inner sep=4pt, fit=(document) (requirements)]{};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(py) (requirements)](code){};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(tex) (bib)](document){};
			\end{scope}
			\node[draw, right=1.5cm of code, fill=green!20] (results) {tmp/results};
			\path[draw, ->] (code) -- (results);
			\node[draw, right=4cm of document, fill=green!20] (pdf) {tmp/ms.pdf};
			\path[draw, ->] (results) |- (pdf);
			\path[draw, ->] (document) -- (pdf);
	\end{tikzpicture}
	\caption{The files of a template that uses the Makefile is depicted in the light gray background and the arrows denote the containerized data flow towards generating the \textit{document}.
	Green indicates temporary cache or generated files, red \textit{\LaTeX\ code} files, blue \textit{results code} files and within the dark gray background, files that are developed by the \textit{author}.}
	\label{fig:templatefiles}
\end{figure}

@@

The Makefile can be called with the following options:
\begin{lstlisting}[language=bash, style=lststyle, caption={Makefile options.}]
make                # Generate the draft (fast) version.
make VERSION=--full # Generate the full (slow) version.
make clean          # Remove the tmp/ directory.
\end{lstlisting}

The \textit{author} may use the Makefile for developing \textit{documents} in the following way:
\begin{enumerate}
	\item clones/downloads the template repository which consists of the Makefile and the set of files as depicted in Fig.\ref{fig:templatefiles},
	\item creates and develops the \textit{results code}.
	\item creates and develops the \textit{\LaTeX\ code}.
	\item executes \textit{make} for fast development iterations and experimentation,
	\item executes \textit{make VERSION=-{}-full} for distribution or publication and
	\item executes \textit{make clean} to restore to a clean state.
\end{enumerate}

@@

A useful \LaTeX\ package for automatically presenting variables from \textit{results} in \textit{\LaTeX\ code} is \textit{datatool} with the following example use:
\begin{lstlisting}[language=TeX, style=lststyle, caption={\LaTeX\ datatool example of loading a file that contains pairs of keys and values (tmp/keys\_values.csv) generated by a \textit{results code} and getting the value of a key named lr.}]
\DTLloaddb{keys_values}{tmp/keys_values.csv}
\DTLfetch{keys_values}{key}{lr}{value}
\end{lstlisting}

@@

Additionally, an \textit{author} can use the \textit{cmp} utility to compare \textit{documents} (generated in a different session, operating system or underlying hardware) byte by byte, and diffoscope for identifying sources of non-determinism as follows:
\begin{lstlisting}[language=bash, style=lststyle, caption={Test draft document version reproducibility. This can also be used as a test script when pushing or pull requesting to a remote repository.}]
# Generate the first draft document version.
make
# Backup the first resulting pdf.
cp tmp/ms.pdf tmp/ms-previous.pdf
# Update the modification timestamp of main.py.
touch main.py
# Generate the second draft document version.
make
# Compare draft document versions byte by byte.
cmp tmp/ms.pdf tmp/ms-previous.pdf
# Compare draft document versions
# in a human readable form.
diffoscope tmp/ms.pdf tmp/ms-previous.pdf
\end{lstlisting}

@@

Making use of the \textit{datatool} package the values of the following variables are not directly referred in the main \textit{.tex} file but they are instead read by an intermediate \textit{.tex} file created by the \textit{results code}: the number of epochs is $\DTLfetch{keys_values}{key}{num_epochs}{value}$, the batch size is $\DTLfetch{keys_values}{key}{batch_size}{value}$ and the learning rate for SGD is $\DTLfetch{keys_values}{key}{lr}{value}$.
Additionally, making use of random seed setting we consistenly get the same figures as shown in Fig.\ref{fig:loss} and Fig.\ref{fig:kernels}.
We also use the \textit{pandas.DataFrame.to\_latex} command which automatically converts a dataframe table to a \LaTeX\ table (as shown in Table~\ref{table:table}):

\begin{figure}[!t]
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/MNIST-loss.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/FashionMNIST-loss.pdf}}
	\\
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/KMNIST-loss.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/QMNIST-loss.pdf}}
	\caption{Comparison of the train and validation loss of MobilenetV2 with ReLU, ReLU6 and SiLU for all datasets.}
	\label{fig:loss}
\end{figure}

\begin{figure}[!t]
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/MNIST-ReLU-kernels.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/FashionMNIST-ReLU-kernels.pdf}}
	\\
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/KMNIST-ReLU-kernels.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/QMNIST-ReLU-kernels.pdf}}
	\caption{The first 25 kernels of the first convolutional layer of MobilenetV2 with ReLU for MNIST (upper-left), FashionMNIST (upper-right), KMNIST (lower-left) and QMNIST (lower-right).}
	\label{fig:kernels}
\end{figure}

\begin{table}[h]
	\centering
	\caption{MobilenetV2 test dataset accuracies.}
	\label{table:table}
	\setlength\tabcolsep{4pt}
	\input{tmp/metrics.tex}
\end{table}

@@

Use cases of the Makefile include:
\begin{itemize}
	\item regression testing and debugging, to ensure that changes to \textit{code} do not alter \textit{results},
	\item common development environment across multiple \textit{authors},
	\item coauthor(s), reviewers, journal editors or other researchers can easily reproduce the \textit{document} with few requirements.
\end{itemize}

>>>>

\begin{document}
\DTLloaddb{keys-values}{tmp/keys-values.csv}

@@

\begin{abstract}
	We propose a Makefile for developing containerized \LaTeX\ technical documents.
	The Makefile allows the author to execute the code that generates variables, tables and figures (results), which are then used during the \LaTeX\ compilation, to produce either the draft (fast) or full (slow) version of the document.
	We also present various utilities that aid in automating the results generation and improve the reproducibility of the document.
	We lastly release an open source repository of this paper that uses the Makefile.
\end{abstract}

@@

Although \textit{literate programming} provides many advantages for demonstrative projects and immediate visual feedback during experimentation, its `dual nature' constrains authors in using specifically designed text editors and/or source code formats thus leading to `app/vendor lock-in' situations.
Moreover, the overlap between natural text and results code makes \textit{literate programming} difficult to debug and version control.

We propose a Makefile for developing containerized \LaTeX\ technical documents.
By using portable technologies that withstood the test of time, we ensure the stability and portability of the proposed Makefile.
We write this paper using a repository that consists of the Makefile and choose Docker as the container technology and the Python programming language for generating the results.

@@

\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[%
			grow via three points={one child at (0.5,-0.7) and two children at (0.5,-0.7) and (0.5,-1.4)},
			edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
			\node[draw](document){document/}
			child{node[draw, anchor=west](makefile){Makefile}}
			child{node[draw, anchor=west, fill=red!20](tex){ms.tex}}
			child{node[draw, anchor=west, fill=red!20](bib){ms.bib}}
			child{node[draw, anchor=west, fill=blue!20](py){main.py}}
			child{node[draw, anchor=west, fill=blue!20](containerfile){Containerfile}}
			child{node[draw, anchor=west, fill=blue!20](requirements){requirements.txt}};
			\begin{scope}[on background layer]
				\node[draw, rounded corners, dashed, fill=gray!20, inner sep=4pt, fit=(document) (requirements)]{};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(py) (requirements)](code){};
				\node[draw, rounded corners, dashed, fill=gray!80, inner sep=2pt, fit=(tex) (bib)](document){};
			\end{scope}
			\node[draw, right=1.5cm of code, fill=green!20] (results) {tmp/results};
			\path[draw, ->] (code) -- (results);
			\node[draw, right=4cm of document, fill=green!20] (pdf) {tmp/ms.pdf};
			\path[draw, ->] (results) |- (pdf);
			\path[draw, ->] (document) -- (pdf);
	\end{tikzpicture}
	\caption{The files of a repository that uses the Makefile is depicted in the light gray background and the arrows denote the containerized data flow towards generating the \textit{document}.
	Green indicates temporary cache or generated files, red \textit{\LaTeX\ code} files, blue \textit{results code} files and within the dark gray background, files that are developed by the \textit{author}.}\label{fig:repositoryfiles}
\end{figure}

@@

The Makefile can be called with the following options:
\begin{lstlisting}[language=bash, style=lststyle, caption={Makefile options.}]
make                # Generate draft (fast) version.\ 
make VERSION=--full # Generate full (slow) version.\ 
make clean          # Remove tmp/ directory.
\end{lstlisting}

The \textit{author} may use the Makefile for developing \textit{documents} in the following way:
\begin{enumerate}
	\item clones/downloads the repository which consists of the Makefile and the set of files as depicted in Fig.\ref{fig:repositoryfiles},
	\item creates and develops the \textit{results code}.
	\item creates and develops the \textit{\LaTeX\ code}.
	\item executes \textit{make} for fast development iterations and experimentation,
	\item executes \textit{make VERSION=-{}-full} for distribution or publication and
	\item executes \textit{make clean} to restore to a clean state.
\end{enumerate}

@@

A useful \LaTeX\ package for automatically presenting variables from \textit{results} in \textit{\LaTeX\ code} is \textit{datatool} with the following example use:
\begin{lstlisting}[language=TeX, style=lststyle, caption={\LaTeX\ datatool example of loading a file that contains pairs of keys and values (tmp/keys-values.csv) generated by a \textit{results code} and getting the value of a key named lr.}]
\DTLloaddb{keys-values}{tmp/keys-values.csv}
\DTLfetch{keys-values}{key}{lr}{value}
\end{lstlisting}

@@

Additionally, an \textit{author} can use the \textit{cmp} utility to compare \textit{documents} (generated in a different session, operating system or underlying hardware) byte by byte, and diffoscope for identifying sources of non-determinism as follows:
\begin{lstlisting}[language=bash, style=lststyle, caption={Test draft document version reproducibility. This can also be used as a test script when pushing or pull requesting to a remote repository.}]
# Generate the first draft document version.\ 
make
# Backup the first resulting pdf.\ 
cp tmp/ms.pdf tmp/ms-previous.pdf
# Update the modification timestamp of main.py.\ 
touch main.py
# Generate the second draft document version.\ 
make
# Compare draft document versions byte by byte.\ 
cmp tmp/ms.pdf tmp/ms-previous.pdf
# Compare draft document versions
# in a human readable form.\ 
diffoscope tmp/ms.pdf tmp/ms-previous.pdf
\end{lstlisting}

@@

Making use of the \textit{datatool} package the values of the following variables are not directly referred in the main \textit{.tex} file but they are instead read by an intermediate \textit{.tex} file created by the \textit{results code}: the number of epochs is $\DTLfetch{keys-values}{key}{num-epochs}{value}$, the batch size is $\DTLfetch{keys-values}{key}{batch-size}{value}$ and the learning rate for SGD is $\DTLfetch{keys-values}{key}{lr}{value}$.
Additionally, making use of random seed setting we consistenly get the same figures as shown in Fig.\ref{fig:loss} and Fig.\ref{fig:kernels}.
We also use the \textit{pandas.DataFrame.to\_latex} command which automatically converts a dataframe table to a \LaTeX\ table (as shown in Table~\ref{table:table}):

\begin{figure}[!t]
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/MNIST-loss.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/FashionMNIST-loss.pdf}}
	\\
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/KMNIST-loss.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/QMNIST-loss.pdf}}
	\caption{Comparison of the train and validation loss of MobilenetV2 with ReLU, ReLU6 and SiLU for all datasets.}\label{fig:loss}
\end{figure}

\begin{figure}[!t]
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/MNIST-ReLU-kernels.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/FashionMNIST-ReLU-kernels.pdf}}
	\\
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/KMNIST-ReLU-kernels.pdf}}
	\subfloat{\includegraphics[width=0.5\columnwidth]{tmp/QMNIST-ReLU-kernels.pdf}}
	\caption{The first 25 kernels of the first convolutional layer of MobilenetV2 with ReLU for MNIST (upper-left), FashionMNIST (upper-right), KMNIST (lower-left) and QMNIST (lower-right).}\label{fig:kernels}
\end{figure}

\begin{table}[ht]
	\centering
	\caption{MobilenetV2 test dataset accuracies.}\label{table:table}
	\setlength\tabcolsep{4pt}
	\input{tmp/metrics.tex}
\end{table}

@@

Use cases of the Makefile include:
\begin{itemize}
	\item regression testing and debugging, to ensure that changes to \textit{code} do not alter \textit{results},
	\item common development environment across multiple \textit{authors},
	\item coauthors, reviewers, journal editors or other researchers can easily reproduce the \textit{document} with few requirements.
\end{itemize}
